Static Analysis by Path Focusing

### 2011-05-20 report/report.tex
loops are now unrolled once before doing the analysis

Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
Abstract Interpretation. It is aimed at computing an upper approximation of the
reachable states of a program containing numerical variables. The set of
possible assignments for the numerical variables is abstracted by convex
polyhedron. This technique discovers invariant linear relations between
the numerical variables at each control point of a program.

@@

The classical union of two convex polyhedra may not be a convex polyhedra.
Therefore, we use the convex hull operation, noted $\sqcup$, for joining
polyhedra. 
If $X_1, X_2$ are two convex polyhedra, $X_1 \sqcup X_2$ is the
smallest convex polyhedron containing both $X_1$ and $X_2$.

@@

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to various reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, a union of convex polyhedra is computed instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

@@

To each point of the control flow graph $\{p_0, p_1,\dots, p_7\}$, we attach a
convex polyhedron, whose dimensions are the numerical variables $x,y$.
For instance, at point $p_1$, the effective set of possible assignments for
$(x,y)$ is the union of the two segments $(0,0) - (51,51)$ and $(51,51) -
(102,0)$, depicted in figure \ref{result}.a. Since this union of segments is not
a convex polyhedra, the best result we can wait for after abstraction is the
polyhedra depicted in figure \ref{result}.b, which is the smallest convex
polyhedra containing the two segments.

@@

b. The best result we can have is the polyhedra in hashed lines.
\end{minipage}
\caption{Best results we can get from the analysis}
\label{result}
\end{figure}

We apply abstract interpretation to this program. Let $X_i$ be the polyhedra
attached to point $p_i$. Initially, $X_i = \perp$ for each $i \in \{1,7\}$, and
$X_0 = \top$. The tabular in figure \ref{iterations} shows the evolution of the
polyhedron during the iterations.

@@

	\subsubsection{Expression of the reachability problem}

	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
	formula expressing the reachability problem defined in \ref{reachability}.

@@

	In the case of path focusing algorithm, we simply conjunct our formula with
	$(i \neq j) \vee (x'_j \neq x_i)$, such that the SMT-solving only proposes
	paths that are not the identity. Actually, it never focuses on such paths
	during the acending sequence, but during narrowing iterations that occur 
	at the end of the algorithm.

@@

@@

	In this section, we show that \emph{Path Focusing} technique performs better
	that \emph{Lookahead} widening on some code taken from real open-source
	programs.

>>>>

Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
Abstract Interpretation. It is aimed at computing an upper approximation of the
reachable states of a program containing numerical variables. The set of
possible assignments for the numerical variables is abstracted by a convex
polyhedron. This technique discovers invariant linear relations between
the numerical variables at each control point of a program.

@@

The classical union of two convex polyhedra may not be a convex polyhedron.
Therefore, we use the convex hull operation, noted $\sqcup$, for joining
polyhedra. 
If $X_1, X_2$ are two convex polyhedra, $X_1 \sqcup X_2$ is the
smallest convex polyhedron containing both $X_1$ and $X_2$.

@@

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to various reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedron including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, a union of convex polyhedra is computed instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

@@

To each point of the control flow graph $\{p_0, p_1,\dots, p_7\}$, we attach a
convex polyhedron, whose dimensions are the numerical variables $x,y$.
For instance, at point $p_1$, the effective set of possible assignments for
$(x,y)$ is the union of the two segments $(0,0) - (51,51)$ and $(51,51) -
(102,0)$, depicted in figure \ref{result}.a. Since this union of segments is not
a convex polyhedron, the best result we can wait for after abstraction is the
polyhedron depicted in figure \ref{result}.b, which is the smallest convex
polyhedron containing the two segments.

@@

b. The best result we can have is the polyhedron in hashed lines.
\end{minipage}
\caption{Best results we can get from the analysis}
\label{result}
\end{figure}

We apply abstract interpretation to this program. Let $X_i$ be the polyhedron
attached to point $p_i$. Initially, $X_i = \perp$ for each $i \in \{1,7\}$, and
$X_0 = \top$. The tabular in figure \ref{iterations} shows the evolution of the
polyhedra during the iterations.

@@

	\subsubsection{Expression of the reachability problem}
	\label{smtformula}
	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
	formula expressing the reachability problem defined in \ref{reachability}.

@@

	In the case of path focusing algorithm, we simply conjunct our formula
	(see \ref{smtformula}) with
	$(i \neq j) \vee (x'_j \neq x_i)$, such that the SMT-solving only proposes
	paths that are not the identity. Actually, it never focuses on such paths
	during the acending sequence, but during narrowing iterations that occur 
	at the end of the algorithm.

@@

\section{Unrolling loops}

In some cases, the number of iterations inside a loop during execution 
may be equals to zero. This happens if the condition of the \emph{while}
statement is false at the first time the head of loop is reached.

In our analyzer, this kind of loops gives imprecise results. In this section, we
propose to unroll loops once in order to avoid it.
This behaviour can happen in very simple programs, such as the one
depicted in Figure \ref{loop}. Indeed, in the case of $n \leq 0$, the condition
$i < n$ is always false.

\begin{figure}[!h]
\begin{minipage}[c]{.29\linewidth}
\begin{C}
i=0;
while (i < n)
	i++;
\end{C}
\end{minipage}
\begin{minipage}[c]{.69\linewidth}
\centering
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.3cm,
                    semithick,font=\footnotesize]

	\node[state] (n0) {$p_0$};
	\node[state] (n1) [below of=n0] {$p_1$};
	\node[state] (n3) [below right of=n1] {$p_3$};
	\node[state] (n4) [below of=n3] {$p_4$};
	\node[state] (n2) [left of=n4,xshift=-1cm] {$p_f$};

  \path [transition] 
		(n0) edge              node {$i \gets 0$} (n1);
  \path [transition] 
        (n1) edge [out=180,in=90]	node [left] {$i \geq n$} (n2);
  \path [transition] 
        (n1) edge			   node [right] {$i < n$} (n3);
  \path [transition] 
		(n3) edge              node {$i \gets i+1$} (n4);
  \path [transition] 
        (n4) edge [out=0, in=0, distance=2.5cm] node {} (n1);

\end{tikzpicture}
\end{minipage}
\caption{If $n \leq 0$, there is no loop iterations.}
\label{loop}
\end{figure}

In this example, our program behaves as follows:
\begin{itemize}
\item At point $p_1$, we first have the polyhedra $i=0$, where $n$ in
unconstrained.
\item The image of this polyhedra after the loop transition is $i \leq n \wedge
i=1$. After a convex hull, we obtain for $p_1$ $0 \leq i \leq 1$.
\item Again, we
compute the image of this polyhedron by the transition, $1 \leq i \leq 2 \wedge
i \leq n$, and after widening we
obtain at $p_1$ the polyhedron $0 \leq i$.
\item Finally, at point $p_f$, we have the polyhedron $0 \leq i \wedge i \geq
n$.
\end{itemize}

Now, after unrolling the loop once, the program is the one in Figure
\ref{loopunroll}. The analysis gives:
\begin{itemize}
\item At point $p_1$, we have $i=0$, then we directly have at point $p_f$ the
polyhedron $i=0 \wedge n \leq 0$.
\item At point $p_4$, we first have $i=1 \wedge n \geq 1$.
\item the image of the polyhedron after the loop transition is $2 \leq n \wedge
i=2$, which gives after a convex hull $ 1 \leq i \leq 2 \wedge i \leq n$.
\item After one iteration and an application of widening, we find at $p_4$ the
polyhedron $1 \leq i \leq n$.
\item Then, at the exit point $p_f'$ of the loop, we have the polyhedron
$i = n \wedge i \geq 1$.
\end{itemize}

\begin{figure}[!h]
\begin{minipage}[c]{.29\linewidth}
\begin{C}
i=0;
if (i < n) {
	i++;
	while (i < n)
		i++;
}
\end{C}
\end{minipage}
\begin{minipage}[c]{.69\linewidth}
\centering
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.3cm,
                    semithick,font=\footnotesize]

	\node[state] (n0) {$p_0$};
	\node[state] (n1) [below of=n0] {$p_1$};
	\node[state] (n3) [below right of=n1] {$p_3$};
	\node[state] (n4) [below of=n3] {$p_4$};
	\node[state] (n2) [left of=n4,xshift=-1cm] {$p_f$};
	\node[state] (n22) [below of=n2,yshift=0.5cm] {$p_f'$};
	\node[state] (n5) [below of=n4] {$p_5$};
	\node[state] (n6) [below of=n5] {$p_6$};

  \path [transition] 
		(n0) edge              node {$i \gets 0$} (n1);
  \path [transition] 
        (n1) edge [out=180,in=90]	node [left] {$i \geq n$} (n2);
  \path [transition] 
        (n1) edge			   node [right] {$i < n$} (n3);
  \path [transition] 
		(n3) edge              node {$i \gets i+1$} (n4);
  \path [transition] 
		(n4) edge              node {$i \geq n$} (n22);
  \path [transition] 
		(n4) edge              node {$i < n$} (n5);
  \path [transition] 
		(n5) edge              node [right] {$i \gets i+1$} (n6);
  \path [transition] 
        (n6) edge [out=0, in=0, distance=2.5cm] node {} (n4);

\end{tikzpicture}
\end{minipage}
\caption{Program of Figure \ref{loop} after unrolling the loop once.}
\label{loopunroll}
\end{figure}

The convex hull of the polyhedron from $p_f$ with the one from $p_f'$ also gives
$0 \leq i \wedge i \geq n$. Then, unrolling the loop does not give a better
solution, since
$p_f$ and $p_f'$ may be the same control point.

However, our technique does a convex hull only if this exit point is in $P_R$. 
If not, we consider independently the paths starting from $p_4$ and the ones
starting from $p_1$, hence the better precision. Indeed, this comes to computing
a disjunctive invariant for the loop, by distinguishing the case where the
number of iterations is zero.

In our analyzer, we unroll every loops once, by using LLVM optimisation
passes such as \emph{loop-unroll}, \emph{loop-rotate} or \emph{loop-simplify}.

@@

	In this section, we show that \emph{Path Focusing} technique performs better
	that \emph{Lookahead widening} on some code taken from real open-source
	programs.
